### 2021-09-06 rad-icfp-2021/rebuttal_POPL.txt

Rebuttal:

We are very grateful to the anonymous reviewers for their detailed feedback. 
If we are given the chance for a revised version, all the minor comments and typos raised by the reviewers have been addressed.

For reviewer2's question:

For reviewer3's questions:

Reviewer 1:

Paper summary
This unary form is like Forth with typed stacks;
A Jacobian amounts to a function from stack-gradients to stack-gradients, 
which can be composed by partial evaluation as long as control flow is static.

Weaknesses
Although the paper is organized so that earlier material motivates later material, insufficient explanation makes understanding much earlier material require understanding later material. For instance:
Understanding Section 3 requires understanding Section 4 (I recommend simply dropping Section 3).
Understanding Section 4 requires understanding Section 6 (I recommend stating the invariant maintained by Figure 8 formally up front).
Understanding Section 5.1 requires understanding Section 5.4 (I recommend merging the material).
Understanding the second line on page 8 (below Figure 4) requires knowing the notation \mathit{Id}_\GammaId 
Γ
​
  (which is introduced in Figure 4, which is not referred to in the text above) and the notation 0_\Gamma0 
Γ
​
  (which is introduced below in Notation 1).
More broadly speaking, understanding this paper requires a prior understanding of reverse-mode AD to an extent that limits its audience severely at POPL.
The paper does not firmly establish the languages that its transformations move among. Examples:

Figure 4, bottom two rows, (x:A[R]^n)(x:A[R] 
n
 ) uses the unbound variable xx
To type-check e_1;e_2e 
1
​
 ;e 
2
​
  in Source UNF, Figure 6 requires T,A \vdash e_2:T,A,BT,A⊢e 
2
​
 :T,A,B, but Figure 9 uses proj and pair after ;;, and the typing rules for proj and pair in Figure 6 do not match T,A \vdash e_2:T,A,BT,A⊢e 
2
​
 :T,A,B.
Figure 9 says "constant seen as a 0-ary operator", which should be stated when defining Source UNF. Typical programming languages only have a finite number of operators.
Page 20 (between Definitions 6.4 and 6.5) interpret a \tilde{;} 
;
~
​
  construct in Source UNF, but there is no such construct in Figure 6.
Although the paper couches semantics and correctness in category theory, it boils down to the sentence "This functor is easily shown to compute reverse-mode derivatives" on page 21, yet the paper never formalizes what it means to "compute reverse-mode derivatives". Ultimately the source language in this paper is so "limited" (Section 2.2) that it doesn't advance our understanding of reverse-mode AD much.

Page 1 "PROPs": what are they?
Page 1 "often complicated heuristics with no guarantees are used": Give concrete examples of those heuristics in a footnote.
Page 2, Example 1.1: The prime ' 
′is used both for dual (w) and for update (y). Only use the prime ' 
′for one of these two purposes; choose a different symbol for the other purpose. Also, rename y1,y2,y3,y4,y5 to Dx1,Dx2,Dx3,Dw1,Dw2 (or use some other letter instead of D).
Refer to Table 1 in the text nearby.
Figure 1 and elsewhere: don't use math mode for italics.
Page 2 "The gradient of each of these expressions with respect to A is:": Clarify whether your transformation turns the 3 lines of code displayed above into the 3 lines of code displayed below.
Page 4, first sentence of Section 2.1, "need both" -> "we need both". The "To" at the beginning of this sentence is confusing: even if our goal is not "to find the gradient ∇f of a function f:R→R compositionally", don't we still "need both f and ∇f when calculating ∇(f;g)"?
Page 4, Section 2.2, define J.
Page 5, "lose compositionality": Clarify what is lost. Compositionality means that the meaning of a complex expression is determined by its structure and the meanings of its constituents.
Page 5, after "by noting that", f_{i+1}(f_i)f 
i+1
​
 (f 
i
​
 ) should be f_{i+1} \circ f_if 
i+1
​
 ∘f 
i
​
 , )()( should be )\circ()∘(, and on the next line, YJ^TfYJ 
T
 f should be Y\circ J^TfY∘J 
T
 f. On the next line, "applying the identity continuation Rn→Rn on the second component" should be "applying the second component to the identity continuation Rn→Rn".

Page 7, "We only need the target language to be second-order.": The target language can probably stay first-order if you treat functions at the meta-level of the transformation.
Page 9, "The idea is that ρ represents the return type of the derivative part" needs to be explained more clearly, with formalization and examples.
Figure 7, where is the "term pairing" e ::= <e,e> construct of target UNF used? I don't see it used in Figure 8.
Page 16, above Theorem 5.4, in the displayed equations, pairs of braces are missing after "max" (perhaps you forgot to escape your braces with backslashes).
Page 19, after "This interpretation extends to", the extra spaces before e are strange, and some e2s seem like they should be e3s. These problems make this paragraph hard to understand.
Page 20 "somewhere in between the syntax" is too vague.
Page 21, the second-to-last paragraph needs to make the following phrases concrete by referring to Example 4.1: "non-Jacobian primitive"? "a tuple"? "highly inefficient"? "projects to the last element"? "non-trivial information of the term"?
Section 7.2 discusses modifications to the composed transformation of Figure 5, but it should be easier to understand (and prove correct and efficient) to modify the decomposed transformations of Section 4 instead.
Page 23, "makes everything harder to optimize" is too broad.
Page 23, define "non-linear usage".
Page 24, "compute the whole gradient of the expression" is unclear. Give examples where we do and examples where we don't want to compute the whole gradient of the expression. It seems that there are common cases of conditionals that your solution does not handle, and that motivates the use of mutable cells allocated at run time.
Show line numbers in draft manuscripts to ease commenting. Number all displayed equations.
Spell check.

Reviewer 2:

Paper summary

Strengths
A unique aspect of the work is to consider performance aspects and to perform complexity analysis where the reverse-mode transformation is still pure.

Weaknesses
The complexity analysis does not consider any parallelism as part of the cost model. Although perhaps reasonable as a first step, the ability to parallelize AD is extremely important in practice.
The work does not include any implementation, neither as an explicit implementation for empirical evaluation, nor a mechanized approach when proving correctness or complexity. Hence, the practical performance implications can only be considered in theory.

Figure 1 shows a descriptive outline of the compilation strategy and an overview of the paper. As shown in the top of the figure and explained in Section 3, there is a direct translation from the source language to the target language. As stated in the middle of page 9, it is neither trivial to understand this approach, nor to prove it correct. And, this is why you describe the second chain via UNF? But, why do you then describe this direct translation in the first place? What does it add, compared to the second chain?
I found the presentation of UNF slightly more complicated than what I think it needs to be. Specifically, when you first introduce the source and target UNF languages (Figures 6 and 7), it is not obvious what the semantics of UNF actually is. It is first in the latter part of the section it becomes clearer, when the transformation from source to source UNF is described. Hence, this section would be more readable if the order is changed, and where the source UNF is introduced together with the transformation from the source language. A more detailed example would also be valuable.
The re-translation step (non-strict transformation UNF^{-1}) is not completely clear. In particular, the reintroduction of variable names, as briefly discussed at the end of page 13.
It would be good if the discussion of how to extend the language further could be elaborated a bit more. For instance, currently the the source language contains product types, but not sum types. How would such an extension affect the approach? Is it direct, and in such a case, why?
On page 24, it says that your approach does not support recursion, which is also clear from the rest of the paper. But, in table 1, the column for "This Paper" shows recursion half filled in. It says in the caption that it means "not supporting general recursion". Is the rational to fill it in for this case, that it can use operations like "reduce" on arrays? It would be good to be clearer regarding this already in the introduction.

Questions for the response period
Why do you present the direct translation (Section 3)? What does it add, compared to performing the translation and differentiation via UNF?

Reviewer 3:

Notable features of the approach include:
The core language studied has no form of recursion, and no conditionals. The paper briefly discusses extensions (in particular with conditionals) at the end, but still banning primitive functions from reals to booleans which generate non-smooth functions.

Weaknesses
Quite a few recent papers propose program transformations for reverse-mode AD of purely functional programs. See for instance [Vákár 2021] which also has arrays, motivates the transformation categorically and validates it via denotational semantics in diffeological spaces. [Mazza and Pagani 2021] do it for PCF, with full support of recursion and conditionals; and there are others.
In contrast, the language considered here is minimalistic. Table 1 seems to claim a partial support for recursion, but it is hard to see what that corresponds to as there is no recursion or iteration in the language. I guess it is map/reduce? The table also claims that conditionals are supported, but this is only briefly sketched in Section 7.2, and this is without primitives R→ B such as >0 making conditionals useful.
The UNF syntax and conversions seems more like a tool than an end in itself, though it does make differentiation more intuitive.
So aside from efficiency, I am not sure which aspect of this work to put forward with respect to existing literature. The paper should also be more upfront about the fact that the main original contribution, with respect to existing work, is a proof of the cheap gradient principle.
Explanations are a bit lacking and assume familiarity which gradients, Jacobians etc.

Comments for author
Section 1.1: the example is hard to follow, because no typing and/or explanation for the components is given, and this appears before the introduction to AD in Section 2.
Section 2.1: this is titled "Rudiments of forward-mode AD and dual numbers", but it only seems to introduce the objective of AD, not forward-mode.
Section 2.2: "The problem with the previous approach shows up when one wants to compute the full gradient of a function R n → R, for a large n. Forward-mode only computes one directional derivative..." -> this sounds weird since we have not introduced forward-mode.
p.4: "Typically ... completely missed" -> this largely ignores the recent literature on the topic.
p.5: the brackets <, > make the spacing really awkward. Please use \langle, \rangle. Plus, pairing of functions is written (f, g) before (in Section 2.1).
p.5: "x′+ = ∂1op(x, y); y′+ = ∂2op(x, y)" -> spacing makes it really hard to parse.
p.5: "We have to carry a continuation and β-reduce a lot of higher-order functions." -> this is a bit imprecise. Why is that a problem?
p.6: "call-by-value" -> sounds a bit strange for a first-order language without calls (nor operational semantics).
p.6: "The types T1 , T2 , terms e1 , e2" -> why not "types T, terms e"? Besides, sometimes indices appear sometimes as subscripts, and sometimes not. This happens inconsistently throughout the whole paper.
Figure 3: the translation also involves partial derivatives of operations, which do not appear here.
Figure 4: "∇Γ e" -> inconsistent, sometimes this is ∇Γ(e), sometimes not.
Figure 4: in the clauses ZerosLike(n) and OnesLike(n), x is unbound.
p.8: reverse-mode AD is not actually defined! Well it is defined in Figure 5, two pages later, and Figure 5 is not referred to before p.12.
Lemma 3.2: but we still do not have the definition!
p.9: "UNF introduces the call-by-value evaluation of the arguments before evaluating the arguments and a good management of the environment" -> ?
Figure 5: missing '(' in the clause for let
p.11: "cos_{R, R}" -> doesn't match the grammar of terms. Should it be "cos_{R, R; 2}"?
Figure 7: the grammar for types implies that one cannot form e.g. [R, R] -> R (as we only add lists of types from target). Is that intended?
Figure 7: what is [A -> B]?
The arrow appears sometimes as ->, sometimes as →
p.13: "Doing UNF followed by UNF^{−1} performs some version of the ANF transformation" -> this sounds weird since later on it is said that UNF^{-1} has no general typing preservation property and should be performed after AD.
Lemma 5.1: "Every variable ... is used exactly once" -> in what sense? Only one occurrence (so not stable under beta-reduction)? Linear typing?
p.15: "This rewriting ... movement of variables" -> this does not make sense before the cost model is defined.
Figure 12: suddenly conditionals appear. Why?
p.17: "The induction hypothesis is then too weak to conclude" -> but you have not said what the IH is.
p.19: "modulo the ηβ-laws" -> undefined, and Source does not have functions. Do you mean ηβ for products?
p.19: "SynSource satisfies the following universal property" -> what about arrays? reduce/map? let?
p.19: "⟨∆n,,[[e2]],[[e2]]⟩; swap;([[e1]])×n" -> why ",,"? Also, spacing is off.
p.20: I was not aware of "concategories". I don't think it is a well-known notion. There should be a reference, including derived notions such as cartesian concategories which are used in the paper without much introduction.
p.20: "explain the UNF and UNF^{−1} transformations." -> font inconsistent.
p.20: "We denote by ... in the context T" -> font for T inconsistent.
Please give the typing of [[-]]!
p.21: Notations \Lambda and \Lambda^{-1} not introduced.
Definition 6.6: "Mutlicategory"
What is \underline{-}? What are u_{\underline{B}_i}? A "++" is split in two "+" around the line.
p.21: "We can interpret Source in C_{Source} as follows" -> is this an interpretation or a functor? A functor from SynSource to C_{Source}?
There are inconsistencies in the font of 'e' below.
The naming convention of categorical structure SynSource/C_{Source} etc is not enlightening. Besides, SynSource is not used beyond p.19.
p.21: "Seen as morphisms in C_{Source}" -> what is seen as a morphism in C_{Source}?
p.22: the diagram is not very clear. Where is AD from Source to Target? Proposition 6.10 corresponds to the commutation of which square?

Questions for the response period
Why does the language not include a primitive to create an array? Doesn't that limit expressivity?

Lemma 5.1: Every variable ... is used exactly once" -> in what sense? Only one occurrence (so not stable under beta-reduction)? Linear typing?

### 2021-09-06 rad-icfp-2021/complexity.tex
starting fixing things after POPL reviews

As can be seen from the examples, $\directD{\rho}{\Gamma}{Y}$ introduces a lot of functions of several arguments.
Following the insight from Section~\ref{subsec:insights}, we do not want to keep all these costly lambda abstractions.
The transformation is designed in such a way that all the lambda abstractions are given arguments,  
and we can use partial evaluation to beta-reduce all these lambda abstractions.

\begin{lemma}
    Let $\Gamma \vdash$!e: A! be a term in Source. 
    Every variable !Y! in $\directD{\rho}{\Gamma}{Y}$(!e!) which is not of ground type is used exactly once.
\end{lemma}

\begin{proof}
    By induction on the judgment of !e!. We inspect each rule in Figure~\ref{fig:direct_diff_macro}.
\end{proof}

From the previous lemma, we see that there is a linear usage of  each continuation $Y$. 
This is one key property which has been ensured by $\UNFSymbol$.
We also note that, apart from the !map2! case, this continuation variable is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
In fact we have $k=2$, except for the !map2! and !reduce!.

@@

\begin{lemma}
    \label{lem:noHO}
    Let $\Gamma \vdash$!e: A! be a term in Source. 
    After the partial evaluation step, $\directD{\rho}{\Gamma}{Y}$(!e!) does not have lambda abstractions.
    Its only variable (bound and free) which does not have a ground type is !Y!.
\end{lemma}

@@

The gives a complexity function $\cost$ valued in $\RR^4$. 
For instance, we have

\begin{tabular}{ll}
    $\cost(*)=(3,0,1,0)$ & $\cost(+)=(3,1,0,0)$\\
    $\cost(c)=(1,0,0,0)$ & $\cost($!sin!$)=(2,0,0,1)$
\end{tabular}

For simplicity, we will not address any parallelism in our cost model. 
So the cost of !map (x.e)! on an array of size $n$ will be $n*\cost$(!e!).
Following Lemma \ref{lem:noHO}, we do not need our cost model to deal with higher-order variables and lambda abstractions.
We can thus restrict our attention to the subset of the Target language which does not contain lambdas, applications or variables which are not of ground type.
The cost function extends compositionally to the restricted Target language. 
It is given in Figure~\ref{fig:costmodel}.

@@

\begin{center}
\begin{tabular}{r c l}
    $\NestA$(!c!) &=& 0 \\
    $\NestA$(!x!) &=& 0 \\
\end{tabular}
    \begin{tabular}{r c l}
    $\NestA$($\pi_i$(!e!)) &=& $\NestA$(!e!)\\
    $\NestA$(!op1 e!) &=& $\NestA$(!e!) \\
\end{tabular}
\end{center}

\begin{center}
\begin{tabular}{r c l}
    $\NestA$(!let x:A = e$_1$ in e$_2$:B!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)  \\ 
    $\NestA$(!< e$_1$, e$_2$ >:AxB!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!) \\ 
    $\NestA$(!e$_1$ op2 e$_2$!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)\\
    $\NestA$(!map2 (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
    $\NestA$(!reduce (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
\end{tabular}
\end{center}

@@

There is a first restriction that prevents our transformation from satisfying the cheap gradient principle.
If we try to show that the cheap gradient principle holds by induction on the term, it fails on the !map2!.
When differentiating !map2!, there are three series of calls to (sub)gradients of !e$_1$!. 
The induction hypothesis is then too weak to conclude. 
The problem comes from the fact that !e$_1$! could itself use !map2!. 
So the constant $K$ can be independant of $n$ and of 
the size of the term if we allow it to be dependent on the level of nestings of !map2!.
A similar phenomenon happens with !reduce!.

@@

In Figure~\ref{fig:optim} we present a of optimizations. 
These optimizations are all obviously valid according to the semantic model. 
As our language is purely functional, we can use these extra optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let-bindings 
and resembles SSA-code \cite{cytron1989efficient}.

Even though our reverse-mode transformation has the right complexity after the partial evaluation step, 
the constant factor is of huge importance in practice. 
The purity of our transformation allows us to make the most out of generic optimizations.  
A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.
In addition, hand-crafted efficient derivatives and more optimizations can easily be added to our language.

>>>>

As can be seen from the examples, $\directD{\rho}{\Gamma}{Y}$ introduces a lot of functions of several arguments.
Following the insight from Section~\ref{subsec:insights}, we do not want to keep all these costly lambda abstractions.
The transformation is designed in such a way that all the lambda abstractions are given arguments,  
and we can use partial evaluation to beta-reduce all these lambda abstractions. 
By inspecting each rule in Figure~\ref{fig:direct_diff_macro}, this allows us to prove by induction on the judgment of !e!:

\begin{lemma}
    Let $\Gamma \vdash$!e: A! be a term in Source. 
    Every variable !Y! in $\directD{\rho}{\Gamma}{Y}$(!e!) which is not of ground type has exactly one occurrence in the term.
\end{lemma}

From the previous lemma, we see that there is a linear usage of each continuation variable $Y$. 
This is one key property ensured by $\UNFSymbol$.
We also note that, apart from the !map2! case, this continuation variable is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
In fact we have $k=2$, except for the !map2! and !reduce!.

@@

\begin{lemma}
    \label{lem:noHO}
    Let $\Gamma \vdash$ !e:A! be a term in Source. 
    After the partial evaluation step, $\directD{\rho}{\Gamma}{Y}$(!e!) does not have lambda abstractions.
    Its only variable (bound and free) which does not have a ground type is !Y!.
\end{lemma}

@@

The gives a complexity function $\cost$ valued in $\RR^4$. 
For primitive operations, we have for instance

\begin{tabular}{ll}
    $\cost$(!*!)=$(3,0,1,0)$ & $\cost($!+!$)=(3,1,0,0)$\\
    $\cost$(!c!)=$(1,0,0,0)$ & $\cost($!sin!$)=(2,0,0,1)$
\end{tabular}

More generally $\cost($!op1!$)=(2,0,0,1)$ for the other unary operations.
For simplicity, we will not address any parallelism in our cost model. 
So the cost of !map (x.e)! on an array of size $n$ will be $n*\cost$(!e!).
Following Lemma \ref{lem:noHO}, we do not need our cost model to deal with higher-order variables and lambda abstractions.
We can thus restrict our attention to the subset of the Target language which does not contain lambdas, applications or variables which are not of ground type.
The cost function extends compositionally to the restricted Target language. 
It is given in Figure~\ref{fig:costmodel}.

@@

\begin{center}
\begin{tabular}{r c l}
    $\NestA$(!c!), $\NestA$(!x!) &=& 0 \\
    $\NestA$($\pi_i$(!e!)), $\NestA$(!op1 e!) &=& $\NestA$(!e!) \\
    $\NestA$(!let x:A = e$_1$ in e$_2$:B!) &=& max($\NestA$(!e$_1$!), $\NestA$(!e$_2$!))  \\ 
    $\NestA$(!< e$_1$, e$_2$ >:AxB!) &=& max($\NestA$(!e$_1$!), $\NestA$(!e$_2$!)) \\ 
    $\NestA$(!e$_1$ op2 e$_2$!) &=& max($\NestA$(!e$_1$!), $\NestA$(!e$_2$!))\\
    $\NestA$(!map2 (x,y.e$_1$) e$_2$ e$_3$!) &=& max(1+$\NestA$(!e$_1$!), $\NestA$(!e$_2$!), $\NestA$(!e$_3$!)) \\
    $\NestA$(!reduce (x,y.e$_1$) e$_2$ e$_3$!) &=& max(1+$\NestA$(!e$_1$!), $\NestA$(!e$_2$!), $\NestA$(!e$_3$!))
\end{tabular}
\end{center}

@@

There is a first restriction that prevents our transformation from satisfying the cheap gradient principle.
If we try to show that the cheap gradient principle holds by induction on the term, it fails on !map2!.
When differentiating !map2!, there are three series of calls to (sub)gradients of !e$_1$!. 
The induction hypothesis is then too weak to conclude. 
The problem comes from the fact that !e$_1$! could itself use !map2!. 
So the constant $K$ can be independant of $n$ and of 
the size of the term if we allow it to be dependent on the level of nestings of !map2!.
A similar phenomenon happens with !reduce!.

@@

In Figure~\ref{fig:optim} we present a list of optimizations. 
These optimizations are all obviously valid according to the semantic model. 
As our language is purely functional, we can use these extra optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let-bindings 
and resembles SSA-code \cite{cytron1989efficient}.

Even though our reverse-mode transformation has the right complexity after the partial evaluation step, 
the constant factor is of huge importance in practice. 
The purity of our transformation allows us to make the most out of generic optimizations.  
In addition, hand-crafted efficient derivatives and more optimizations can easily be added to our language.

### 2021-09-06 rad-icfp-2021/correctness.tex
Indices fix
starting fixing things after POPL reviews

\begin{tabular}{l c l}
    $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& $\pi_T \widehat{+}(0,\pi_A,0)$ \\
    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& $id_T\concatcomp map2_{T;x,y.e}$ \\
    $\seml$!e$_1$!$\icomp$!e$_2$!$\semr$ &=& $\Lambda$($id_\Gamma\concatcomp\seml$!e2!$\semr$; $\Lambda^{-1}$($\seml$!e1!$\semr$)) \\
    $\seml$!J!$^T$!reduce!$_{T;x,y.e_1;e_2}\semr$  &=& $id_T\concatcomp reduce_{T;x,y.e_1;e_2}$ \\
    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T_1}\concatcomp\widehat{0}_{T_2}\concatcomp id_{T_3}$ 
    \end{tabular}
    \begin{tabular}{l c l}
    $\seml$!J!$^T$!op!$_{T;n} \semr$ &=& $id_T\concatcomp J^Top_n$ \\
    $\seml$!<e$_1$, e$_2$>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
    $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& $id_T\concatcomp$!<$\pi_1,\pi_2$>! \\
 \end{tabular}\\
 where !T,A! $\vdash$ (0,$\pi_A$,0): !T! and the projection $\pi_A$ lands in the $i$-th element of the list !T!.

@@

This interpretation is in essence $\UNFSymbol^{-1}$. 
The difference is that to preserve typing, the semantic $\UNFSymbol^{-1}$
sends a non-Jacobian primitive to a tuple, as in the Example \ref{exm:unf}. 
This is highly inefficient, and the syntactic $\UNFSymbol^{-1}$ 
additionally projects to the last element, where the non-trivial information of the term is.

What remains to explain now is the reverse mode transformation between Source UNF and Target UNF.
We construct a functor $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$. 
This functor is easily shown to compute reverse-mode derivatives.
Because terms of Source UNF in $\ConcatS$, we observe the effect of $\Dsynrevsymbol_{\rho}$ on them
and show that it matches the syntactic $\Dsynrevsymbol_{\rho}$ 
from Section~\ref{sub:Simple reverse mode transformation}.

\begin{definition}[$\Dsynrevsymbol_{\rho}$ as a lax functor]
    We define $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$ as follows.\\
    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
    !$\Dsynrevsymbol_{\rho}$(op1)=<op1,$\pi_{last}\circ$ J$^T$op1>!,\\ 
    !$\Dsynrevsymbol_{\rho}$(op2)=<op2,$\pi_{last}\circ$ J$^T$op2>!,\\
    !$\Dsynrevsymbol_{\rho}$(map2 (x,y.e))=<map2 (x,y.e),$\pi_{last}\circ$ J$^T$map2 (x,y.e)>!,\\
    !$\Dsynrevsymbol_{\rho}$(reduce$_{x,y.e_1,e_2}$)=<reduce$_{x,y.e_1,e_2}$,$\pi_{last}\circ$ J$^T$reduce$_{x,y.e_1,e_2}$>!, 
    where $\pi_{last}$ is the projection to the last element (the continuation variable).
    It naturally extends on sequential composition. 
    It does not automatically extend to a multi-functor between concategories because \\
    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$]) $\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp\Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.\\
    Still, there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.
    Internally, as a lambda term, it is given by \\
    $(x_1,\ldots,x_k,Y_1,(x_{k+1},\ldots,x_n,Y_s)\mapsto (x_1,\ldots,x_n,\lambda(y_1,\ldots,y_n)\to Y_1(y_1,\ldots,y_k)\widehat{+}Y_2(y_{k+1},\ldots,y_n)$.
    Here, $\widehat{+}$ is the reverse-derivative of the copy map which is known to be fanout. 
    It is not surprising to see it appear as $\Dsynrevsymbol_{\rho}$ is a semantic functor, and it does not need to be efficient. 
\end{definition}

>>>>

\begin{tabular}{l c l}
    $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& $\pi_T \widehat{+}(0,\pi_A,0)$ \\
    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& $id_T\concatcomp map2_{T;x,y.e}$ \\
    $\seml$!e$_1$!$\icomp$!e$_2$!$\semr$ &=& $\Lambda$($id_\Gamma\concatcomp\seml$!e$_2$!$\semr$; $\Lambda^{-1}$($\seml$!e$_1$!$\semr$)) \\
    $\seml$!J!$^T$!reduce!$_{T;x,y.e_1;e_2}\semr$  &=& $id_T\concatcomp reduce_{T;x,y.e_1;e_2}$ \\
    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T_1}\concatcomp\widehat{0}_{T_2}\concatcomp id_{T_3}$ 
    \end{tabular}
    \begin{tabular}{l c l}
    $\seml$!J!$^T$!op!$_{T;n} \semr$ &=& $id_T\concatcomp J^Top_n$ \\
    $\seml$!<e$_1$, e$_2$>!$\semr$  &=& !<!$\seml$!e$_1$!$\semr$, $\seml$!e$_2$!$\semr$!>! \\
    $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& $id_T\concatcomp$!<$\pi_1,\pi_2$>! \\
 \end{tabular}\\
 where !T,A! $\vdash$ (0,$\pi_A$,0): !T! and the projection $\pi_A$ lands in the $i$-th element of the list !T!.

@@

This interpretation is in essence $\UNFSymbol^{-1}$. 
The difference is that to preserve typing, the semantic $\UNFSymbol^{-1}$
sends a non-Jacobian primitive to a tuple, as in Example \ref{exm:unf}. 
This is highly inefficient, and the syntactic $\UNFSymbol^{-1}$ 
additionally projects to the last element, where the non-trivial information of the term is.

What remains to explain now is the reverse mode transformation between Source UNF and Target UNF.
We construct a functor $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$. 
This functor computes reverse-mode derivatives. \MH{define more precisely}
Because terms of Source UNF are interpreted in $\ConcatS$, we observe the effect of $\Dsynrevsymbol_{\rho}$ on them
and show that it matches the syntactic $\Dsynrevsymbol_{\rho}$ 
from Section~\ref{sub:Simple reverse mode transformation}.

\begin{definition}[$\Dsynrevsymbol_{\rho}$ as a lax functor]
    We define $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$ as follows.\\
    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
    !$\Dsynrevsymbol_{\rho}$(op1)=<op1,$\pi_{last}\circ$ J$^T$op1>!,\\ 
    !$\Dsynrevsymbol_{\rho}$(op2)=<op2,$\pi_{last}\circ$ J$^T$op2>!,\\
    !$\Dsynrevsymbol_{\rho}$(map2 (x,y.e))=<map2 (x,y.e),$\pi_{last}\circ$ J$^T$map2 (x,y.e)>!,\\
    !$\Dsynrevsymbol_{\rho}$(reduce$_{x,y.e_1,e_2}$)=<reduce$_{x,y.e_1,e_2}$,$\pi_{last}\circ$ J$^T$reduce$_{x,y.e_1,e_2}$>!, 
    where $\pi_{last}$ is the projection to the last element (the continuation variable).
    It naturally extends to sequential composition. 
    It does not automatically extend to a multi-functor \MH{define} between concategories because \\
    !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$]) $\neq$ $\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp\Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])!.\\
    Still, there is a map !$\Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$])$\concatcomp \Dsynrevsymbol_{\rho}$([A$_{k+1}$,$\ldots$,A$_n$])$\to \Dsynrevsymbol_{\rho}$([A$_1$,$\ldots$,A$_k$]$\concatcomp$[A$_{k+1}$,$\ldots$,A$_n$])!.
    Internally, as a lambda term, it is given by \\
    $(x_1,\ldots,x_k,Y_1,(x_{k+1},\ldots,x_n,Y_s)\mapsto (x_1,\ldots,x_n,\lambda(y_1,\ldots,y_n)\to Y_1(y_1,\ldots,y_k)\widehat{+}Y_2(y_{k+1},\ldots,y_n)$.
    Here, $\widehat{+}$ is the reverse-derivative of the copy map, which is known to be fanout \MH{define}. 
    It is not surprising to see it appear as $\Dsynrevsymbol_{\rho}$ is a semantic functor, and it does not need to be efficient. 
\end{definition}

### 2021-09-06 rad-icfp-2021/generalization.tex
starting fixing things after POPL reviews

We can now compute the reverse-mode transformation as follows.

@@

\begin{tabular}{l}
    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) \\
    !let b=e$_1$ in <if b then e$_2$ else e$_3$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
    !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e$_2$)(x$_1$,$\ldots$,x$_n$,z)+(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e$_3$)(x$_1$,$\ldots$,x$_n$,z))>!
\end{tabular}

@@

\subsection{Recipe for Adding More Constructs}
\label{sub:lift_recipe}
To add more constructs to our language, 
there is no need to go through UNF; most of the backend of our work can be used as a black box.
The non-smooth variables of the operator (typically booleans or integers) should be considered as external to the language, similarly to the way we treat the index $n$ of arrays.

>>>>

@@

\begin{tabular}{l}
    $\directD{\rho}{\Gamma}{Y}$(!if e$_1$ then e$_2$ else e$_3$!) = \\
    !let b=e$_1$ in <if b then e$_2$ else e$_3$, fun (x$_1$,$\ldots$,x$_n$,z) ->! \\
    !Y(b*!$\directD{\rho}{\Gamma}{Y}$!(e$_2$)(x$_1$,$\ldots$,x$_n$,z)+(1-b)*!$\directD{\rho}{\Gamma}{Y}$!(e$_3$)(x$_1$,$\ldots$,x$_n$,z))>!
\end{tabular}

@@

\subsection{Recipe for Adding More Constructs}
\label{sub:lift_recipe}
To add more constructs to our language, 
there is no need to go through UNF; most of the backend of our work can be used as a black box.
Non-smooth variables (typically booleans or integers) should be considered as external to the language, similarly to the way we treat indices $n$ of arrays.
